---
title: "GLMER_LongitudinalData"
author: "Pietro Mingardi"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
rm(list= ls())
options(scipen = 999)

library(readxl)
library(tidyverse)
library(lme4)
library(car)
library(emmeans)
library(ggpubr)
library(dplyr)
```

## Introduction

This dataset contains information from an experimental study in which foreign students with different native languages were tested in a psycholinguistic task. We aim to analyze whether their performance improved from the first time they performed the task (Time 0) to the second time (Time 1), six months later. We also aim to determine whether any observed improvements were mediated by their native language. 

```{r import data}
#import dataset
df = read_excel("Df.xlsx")

df$Word = as.factor(df$Word)
df$puID = as.factor(df$puID)
df$time = as.factor(df$time)
df$LanguageFamily = as.factor(df$LanguageFamily)

summary(df)

#check numerosity for each language family
numerosity = df %>%
  group_by(LanguageFamily) %>%
  summarise("num" = n_distinct(puID))

numerosity
```
The sample size is relatively small. Moreover, three participants reported Italian as their mother tongue; these students will therefore be excluded from the sample. Given that our primary interest is to test whether different native languages affect performance, and considering that the small number of participants within each Language Family would likely produce unstable effect estimates, we decided to aggregate participants based on the linguistic roots of their native language. French and Spanish will be grouped as Romance languages, while German and Austrian will be grouped as Germanic languages. Belgian participants will be classified according to their native language (four are French-speaking and one is German-speaking).
```{r data quality}
#recodify languages
df_new <- df %>%
  filter(LanguageFamily != "Italian") %>%   # elimina italiani
  mutate(LangGroup = case_when(
    LanguageFamily %in% c("French", "Spanish") ~ "Romance",
    LanguageFamily %in% c("German", "Austrian") ~ "Germanic", 
    LanguageFamily == "Belgian" & puID != "p3" ~ "Romance",
    LanguageFamily == "Belgian" & puID == "p3" ~ "Germanic"
  ), LangGroup = factor(LangGroup))

df_new %>%
  group_by(LangGroup) %>%
  summarise("n" = n_distinct(puID))

#summary(df_new) #check new dataframe

df_new$LanguageFamily <- droplevels(df_new$LanguageFamily) #remove italian level


#plot performance differences between time0 and time1 for the two different LanguaGroup created
descr = df_new %>%
  group_by(puID, LangGroup, time) %>%
  summarise("meanCorrect" = mean(Correct), .groups = "drop") 

ggplot(descr, aes(x = LangGroup, y = meanCorrect, fill = time))+
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(
    x = "Language Group",
    y = "Proportion correct (per subject)",
    fill = "time"
  ) +
  ylim(0, 1) +
  theme_minimal()


#check NA's
colSums(is.na(df_new))

```
The plot indicates greater variability in performance for the Romance group, particularly at Time 0. In contrast, the Germanic group exhibits more homogeneous performance at both measurement points. Moreover, both groups appear to show an improvement from Time 0 to Time 1, as reflected by the higher median scores at the second measurement. The increase seems more pronounced for the Germanic group, although this pattern should be confirmed through statistical analysis.

## Model fitting
```{r model fitting}
#define costrast
contrasts(df_new$time) <-  matrix(c(-0.5, 0.5), ncol = 1)
contrasts(df_new$LangGroup) <-  matrix(c(-0.5, 0.5), ncol = 1)

model_DEF = glmer(Correct ~ time*LangGroup + (1 + time|puID) + (1|Word), family = "binomial", 
                data= df_new, control=glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5)))

#check singularity of the random structure
isSingular(model_DEF)
```
The random-effects structure of the model was initially specified to include random slopes for Times by the grouping variable ParticipantsID and random slopes for Time by the grouping variable Word. However, this structure gieven a singular fit warning. Therefore, we simplified the random structure stepwise until a model without convergence or singularity warnings was obtained. Specifically, we first removed the correlation between the random slopes of Time and the random intercept for the grouping variable Word. The final random-effects structure reported here corresponds to the most complex model that converged without warnings.

```{r coefficients estimates}
summary(model_DEF) #let's have a look at the coefficients estimated by the model
```
To evaluate whether the two-way interaction was statistically significant, we compared the full model (including the interaction term) with a reduced model including only the main effects. Model comparison was performed using a likelihood ratio test based on the Chi-square statistic.
```{r check interaction}

model_base = glmer(Correct ~ time + LangGroup + (1 + time|puID) + (1|Word), family = "binomial", 
                     data= df_new, control=glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5)))

anova(model_base, model_DEF, test = "Chisq" )
```
The interaction was statistically significant. To further investigate our hypotheses, we conducted post hoc tests.
```{r pairwise comparisons}
emm  = emmeans(model_DEF, ~time:LangGroup, type = "response")
emm


Germanic_time0 = c(1, 0, 0, 0)
Germanic_time1 = c(0, 1, 0, 0)
Romance_time0 = c(0, 0, 1, 0)
Romance_time1 = c(0, 0, 0, 1)

#wihtout correction
uncorrect_p = contrast(emm,  method = list("Germanic_time1 - Germanic_time0" = Germanic_time1 - Germanic_time0,
                             "Romance_time1 - Romance_time0" = Romance_time1 - Romance_time0,
                             "Germanic_time0 - Romance_time0" = Germanic_time0 - Romance_time0,
                             "Germanic_time1 - Romance_time1" = Germanic_time1 - Romance_time1
                             ))

uncorrect_p

#with p-value correction for multiple comparisons
correct_p = contrast(emm,  method = list("Germanic_time1 - Germanic_time0" = Germanic_time1 - Germanic_time0,
                             "Romance_time1 - Romance_time0" = Romance_time1 - Romance_time0,
                             "Germanic_time0 - Romance_time0" = Germanic_time0 - Romance_time0,
                             "Germanic_time1 - Romance_time1" = Germanic_time1 - Romance_time1
                             ), adjust ="holm")

correct_p
```
# Intepretation

A likelihood ratio test revealed a significant Time*LangGroup interaction, X2(1) = 5.313, p = 0.021.
Post hoc comparisons showed that the Germanic group significantly improved from Time 0 to Time 1 (odds ratio = 2.203, SE = 0.441, z = 4.236, p < 0.001). In contrast, the improvement observed in the Romance group was not statistically significant.
The two groups did not significantly differ at Time 0. At Time 1, the difference—favoring the Germanic group—approached significance after correction for multiple comparisons using the Holm–Bonferroni method (Germanic vs. Romance at Time 1: odds.ratio = 1.655, S.E. = 0.390, z.ratio = 2.138, p = 0.098).

# Check model robustness
```{r model robustness}
#check if single subject affects beta estimations and p-values of the two-way interaction detected
subjects <- unique(df_new$puID)

betas <- lapply(subjects, function(s) {
  fit <- update(model_DEF, data = subset(df_new, puID != s))
  fixef(fit)
})

betas_mat <- do.call(rbind, betas)
apply(betas_mat, 2, range)


lrt_p <- sapply(subjects, function(s) {
       d <- subset(df_new, puID != s)
       m0 <- update(model_base, data = d)
       m1 <- update(model_DEF, data = d)
       anova(m0, m1, test="Chisq")$`Pr(>Chisq)`[2]
   })
range(lrt_p)
```
Leave-one-out analyses confirmed that the results and model estimates were not substantially affected by the exclusion of any single participant.

# Plot of model estimations
```{r plot}
emm_df <- as.data.frame(emm) 

ggplot(emm_df, aes(x = time, y = prob, 
                   color = LangGroup, 
                   group = LangGroup)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL),
                width = 0.1) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "Time",
    y = "Predicted probability",
    color = "Language Group"
  ) +
  theme_minimal()
```